# Markdown & Tidyverse {#seminar2}

## Markdown

Last week we wrote or codes on the `Source` pane, and if you save the codes and revisit the file with your file explorer, you can see that the extension of the file is `.R`. You can open the file with any other text editor (like Notepad) and see the codes. In a `.R` file you can write your codes and comments, and if you hit `ctrl + shift + enter` then all lines will be evaluated. Most of the times the goal to use `.R` files is this, we want to reuse the codes frequently (like downloading data from a website).

Another possible extension for your files is the `.Rmd`, which stands for *R MarkDown*. In this file format you can combine text, R codes and their output into one single document. from now, We will use `RMarkdown` during the class. Please visit the following website to find useful examples: <https://rmarkdown.rstudio.com/articles_intro.html>

## Introduction to the tidyverse

We will work with the tidyverse today. You can simply install it from the CRAN (go to `Packages` pane and click the install button). But tidyverse is not a simple package. "The tidyverse is a set of packages that work in harmony because they share common data representations [...]"

`Tidyverse` contains the most important packages that you're likely to use in everyday data analyses:

-   [ggplot2](https://ggplot2.tidyverse.org/), for data visualisation.

-   [dplyr](https://dplyr.tidyverse.org/), for data manipulation.

-   [tidyr](https://tidyr.tidyverse.org/), for data tidying.

-   [readr](https://readr.tidyverse.org/), for data import.

-   [purrr](https://purrr.tidyverse.org/), for functional programming.

-   [tibble](https://tibble.tidyverse.org/), for tibbles, a modern re-imagining of data frames.

-   [stringr](https://github.com/tidyverse/stringr), for strings.

-   [forcats](https://github.com/tidyverse/forcats), for factors.

After you installed the packages, you can load in the core packages with \`library\` command. You will receive a warning message, but do not worry, this is fine.

```{r message=T, warning=T}
library(tidyverse)
```

## The `%>%` operator

This operator is written `ctrl + shift + m`, it is denoted as `%>%`, but we pronounce as **pipe**. To understand its relevance, you should think about it like *... then*. This opreator forward the value of the previous expression into the next function as the first unspecified input. Lets see an example:

We first generate a numerical vector as trajectories of standard normal distribution.

```{r fig.cap="Histogram of standard normal distribution"}
norm_sample <- rnorm(n = 1000, mean = 0, sd = 1) # standard normal
```

**AND THEN** we visualize its distribution with a histogram.

```{r fig.cap="Histogram of standard normal distribution"}
hist(norm_sample)
```

Now we had to assign the `norm_sample` object to use only for a single graph. If you are working on a project it will be confusing to have tons of one time used objects in your `environment` (Like `DataFrameAfterCleaningStep1`, `DataFrameAfterCleaningStep2`, `DataFrameAfterCleaningStep3`, etc.).

With the pipe operator we can embed several steps into one single workflow. This way we do not have to assign the `norm_sample` object. We simple generate the random values AND THEN draw their distribution.

```{r fig.cap="Histogram of standard normal distribution using the pipe"}
rnorm(n = 1000, mean = 0, sd = 1) %>% 
  hist()
```

You may see that the title of plot has changed (probably the graph also changed a bit, since chances are low that we generate exactly the 1000 random values). This is because pipe and many other tidyverse function works with a *lambda like* function framework. This means that you can refer to the input value (only if you have only one single input) with `.`.

Example:

```{r}
(2 + 2) %>% 
  {. * .}
```

The result is 16, since 2 + 2 = 4, and 4 \* 4 equals 16.

The pipe may seem unrelevant first, but this is one of the most powerful tool in R if you have complex data manipulating steps. We will cover the core functions of `dplyr` in this chapter, which appear in the following video (37:30 - 38:30), but when you are familier with them, you will see the motivation behind the pipe.

<iframe width="560" height="315" src="https://www.youtube.com/embed/8SGif63VW6E?start=2249" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>

## Tibble, count, filter, select, arrange

### Tibble

Let's import a dataset from the [OECD webpage](https://data.oecd.org/pop/fertility-rates.htm#indicator-chart). Download the data in csv format and paste it into your working directory.

```{r }
fertility_df <- read_csv("DP_LIVE_22092021161631568.csv")
```

`read_csv` is importad from the `readr` package, thus shares its output is adequat to the tidyverse principles. `fertility_df` is a data.frame, but it is a `tibble`, which means that it will be printed nicely on the console: only the first 10 rows appears, and only the amount of columns that can be printed without a linebreak (similar to pandas dataframe in python).

```{r}
fertility_df
```

### Count

With the `count` function you can simple *count* the number of appearancesof the levels of a given column. Example:

```{r}
count(fertility_df, LOCATION)
```

Now we can see that we have 60 observation from ARG. If you are interested in for which country we have the most datapoints, you can use the `sort = TRUE` option.

```{r}
count(fertility_df, LOCATION, sort = TRUE)
```

You can also use the pipe here[^It would not be a good solution to assign a new dataframe to each count table]

```{r}
fertility_df %>% # %>% = the 1st argument of count function is the fertility_df data.frame
  count(LOCATION)
```

You may specifiy multiple columns, this way the frequency of all value combinaton appears.

```{r}
fertility_df %>% 
  count(LOCATION, TIME, sort = T)
```

Now we see that we have only one observation for each country at a time, so this data frame was originally tidy (this is usually not the case). Let's see for example the Q GDP data from the [OECD webpage](https://data.oecd.org/gdp/quarterly-gdp.htm).

```{r}
gdp_df <- read_csv("DP_LIVE_22092021163414835.csv")
```

Using the `count` function, we may find that `MEASURE`, `SUBJECT` & `FREQUENCY` columns have multiple levels.

```{r}
gdp_df %>% 
  count(MEASURE, SUBJECT, FREQUENCY)
```

Let's say we are interested only in the observations where the `SUBJECT` is equal to TOT (total), `MEASURE` is equal to PC_CHGPY (percentage change to previous value) and the `FREQUENCY` is not equal to A (onyl quarterly date).

```{r}
gdp_df %>% 
  filter(SUBJECT == "TOT" & MEASURE == "PC_CHGPY" & FREQUENCY != "A") %>% 
  mutate(TIME = lubridate::yq(TIME)) %>% 
  filter(TIME == max(TIME)) %>% 
  arrange(desc(Value)) %>% 
  select(geo = LOCATION, gdp_change = Value)
```

`filter`: filter the rows, where the condition(s) are TRUE `select`: select the specified columns `mutate`: specify a new column or change an existing one

```{r}
gdp_df %>% 
  select(1)

gdp_df %>% 
  select(-1)

gdp_df %>% 
  select(1:2)

gdp_df %>% 
  select(LOCATION:TIME)

gdp_df %>% 
  select(TIME, LOCATION, everything())
```

```{r}
gdp_df %>% 
  mutate(g = 1 + Value/100)

gdp_df %>% 
  mutate(Value = 1 + Value/100)

gdp_df %>% 
  filter(FREQUENCY != "A") %>% 
  mutate(
    TIME = lubridate::yq(TIME),
    Value = 1 + Value/100
  )
```

`summary`

```{r}
eurostat::search_eurostat("birth")
```

```{r}
livebirth_eu_df <- eurostat::get_eurostat("demo_r_fagec")
```

```{r}
livebirth_eu_df %>% 
  filter(str_length(geo) != 2) %>% 
  filter(age != "TOTAL" & time == "2019-01-01") %>% 
  filter(!(age %in% c("UNK", "Y_GE45", "Y_GE48", "Y_GE50", "Y_LT16"))) %>% 
  group_by(age) %>% 
  summarise(values = sum(values))
```

```{r}
x <- c(100, 107, 105, 110)
chain_index(x)
```

```{r}
chain_index <- function(x) {
  scales::percent(x/lag(x)-1, accuracy = .01)
}
```

## Pivot longer/wider

```{r}
fertility_df %>% 
  select(geo = LOCATION, time = TIME, fertility = Value) %>% 
  pivot_wider(names_from = "geo", values_from = "fertility") %>% 
  mutate_at(-1, chain_index) %>% 
  pivot_longer(-1, names_to = "geo", values_to = "fertility")
```

```{r}
fertility_df %>% 
  select(geo = LOCATION, time = TIME, fertility = Value) %>% 
  arrange(time) %>% 
  group_by(geo) %>% 
  group_modify(~ mutate(.x, fertility_growth = chain_index(fertility)), .keep = F)
  
```

ggplot(data, mapping) + geom

```{r}
df_to_plot <- fertility_df %>% 
  filter(TIME == 2018) %>% 
  select(geo = LOCATION, fertility = Value)
```

```{r}
ggplot(data = df_to_plot, mapping = aes(x = fertility, y = geo)) + 
  geom_col()
```

```{r}
df_to_plot2 <-fertility_df %>% 
  filter(TIME ==2018) %>% 
  select(geo = LOCATION, fertility = Value) %>% 
  mutate(continent = countrycode::countrycode(sourcevar = geo, origin  = "iso3c",
                                              destination = "continent"),
         geo = fct_reorder(geo, fertility)) %>%
  filter(continent == "Europe")
```

```{r}
ggplot(data = df_to_plot2, mapping = aes(x = fertility, y = geo)) + 
  geom_col()
```

```{r}
fertility_df %>% 
  filter(TIME ==2018) %>% 
  select(geo = LOCATION, fertility = Value) %>% 
  mutate(continent = countrycode::countrycode(sourcevar = geo, origin  = "iso3c",
                                              destination = "continent"),
         geo = fct_reorder(geo, fertility)) %>%
  filter(continent == "Europe") %>% 
  ggplot(mapping = aes(x = fertility, y = geo)) + 
  geom_col(fill = "green", color = "black")
```
